Case-based Arabic Morphological Analysis

### 2018-04-12 tex/merf/TALMajor/sarf.tex
updated related work
changes to answer, colors, and some reviewers addressed

%~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
%{\bf Morphological analyzer (Sarf).~}
Morphological analysis is key to Arabic NLP~\cite{arabicmorph} 
due to the exceptional degree of ambiguity in writing, 
the rich morphology, and the complex word derivation system. 
Short vowels, also known as diacritics, are typically omitted in Arabic text
and inferred by readers~\cite{habash2006arabic}. 
For example, the word \RL{bn}%
can be interpreted as \RL{bon} {\tt ``coffee''} with a {\em damma} diacritic on the 
letter \utfrl{بـ} or 
\vocalize \RL{bin} (son of) with a 
{\em kasra} diacritic on the letter \utfrl{بـ}.

@@

We denote by 
\ss,
\pp,
\xx,
\PP,
\GG, and 
\AC, the set of 
all stems,
prefixes,
suffixes,
POS,
gloss, 
and user defined category tags, respectively. 
Let $T=\langle t_1,t_2,\ldots,t_M\rangle$ be a set of Arabic words denoting the 
text documents. 
\framework uses Sarf to compute a set of morphological solutions $M(t)=\{m_1,m_2,\ldots,m_N\}$
for each word $t\in T$. 
Each morphological solution $m\in M(t)$ is a tuple of the form 
$\langle p,s,x,P,G,C\rangle\in \pp \times  \ss\times \xx\times \PP\times\GG \times \AC$.

>>>>

%~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
%{\bf Morphological analyzer (Sarf).~}
\major{
Morphological analysis is key to Arabic 
NLP
due its to the exceptional degree of ambiguity in writing, 
the rich morphology, and the complex word derivation 
system~\cite{arabicmorph,shahrour2016camelparser,pasha2014madamira}.
}
Short vowels, also known as diacritics, are typically omitted in Arabic text
and inferred by readers~\cite{habash2006arabic}. 
For example, the word \RL{bn}%
can be interpreted as \RL{bon} {\tt ``coffee''} with a {\em damma} diacritic on the 
letter \utfrl{بـ} or 
\vocalize \RL{bin} (son of) with a 
{\em kasra} diacritic on the letter \utfrl{بـ}.

@@

We denote by 
\ss,
\pp,
\xx,
\PP,
\GG, and 
\AC, the set of 
all stems,
prefixes,
suffixes,
POS,
gloss, 
and user defined category tags, respectively. 
Let $T=\langle t_1,t_2,\ldots,t_M\rangle$ be a set of Arabic words denoting the 
text documents. 
\framework uses Sarf to compute a set of morphological solutions $M(t)=\{m_1,m_2,\ldots,m_N\}$
for each word $t\in T$. 
%
\major{
Each morphological solution $m\in M(t)$ is a tuple of the form 
$\langle p,s,x,P,G,C\rangle \in \pp \times \ss \times \xx \times \PP \times \GG \times \AC$ 
where $p=p_1 \ldots p_{|p|}$, $x=x_1 \ldots x_{|x|}$, 
$P=P_{p_1}\ldots P_{p_{|p|}} P_s P_{x_1}\ldots P_{x_{|x|}}$,
$G=G_{p_1}\ldots G_{p_{|p|}} G_s G_{x_1}\ldots G_{x_{|x|}}$, and
$C=C_{p_1} \ldots C_{p_{|p|}} C_s C_{x_1}\ldots C_{x_{|x|}}$. 
$P_{p_i}, G_{p_i},$ and $C_{p_i},1\le i \le |p|$ are the POS, Gloss and 
category tags of prefix $p_i$. 
$P_{x_j}, G_{x_j},$ and $C_{x_j},1\le j \le |x|$ are the POS, Gloss and 
category tags of suffix $x_i$. 
$P_s, G_s,$ and $C_s$ are 
$P_s, G_s,$ and $C_s$ are the POS, Gloss and 
category tags of stem $s$. 
%
Intuitively, 
$p,x,P,G$ and $C$ are concatenations of prefix, suffix, POS, Gloss and
Category values, respectively.
}

### 2018-04-12 tex/merf/TALMajor/tool.tex
changes to answer, colors, and some reviewers addressed

\subsection{The extended synonymy feature $Syn^k$}

The motivation for $Syn^k$ is to construct a light Arabic ontology based on the lexicon of Sarf.
The sets $E, A,$ and $L$ denote all English words, Arabic words, 
and Arabic lexicon words, respectively.
Recall that $\GG$ and $\ss$ denote the set of glosses and stems in the morphological analyzer, respectively.
We have $\GG \subset E$ and $\ss \subset L \subset A$. 
Function $\alpha: \ss \rightarrow 2^{\GG}$ maps Arabic stems to 
subsets of related English glosses. %; e.g. $g_{s} = \alpha(s)\subset 2^{\GG}$.
Function $\gamma: L \rightarrow 2^{\ss}$ maps Arabic lexicon words to subsets 
of relevant Arabic stems. %; e.g. $s_{l} = \gamma(l)\subset 2^{\ss}$.

Given a word $w\in L$, 
$Sy(w)=\{u\mid u\in \ss \land\exists s\in \gamma(w)\land~\alpha(u)\cap\alpha(s)\neq\emptyset\}$
is the set of Arabic stems 
directly related to $w$ through the gloss map.

Let $Sy^{i}(w)$ denote stems related to $w$ using the gloss map of order $i$ recursively such that
$Sy^{1}(w) = Sy(w)$ and
$Sy^{i+1}(w)=\{u\mid u\in S\land\exists s\in Sy^{i}(w)\land~\alpha(u)\cap\alpha(s)\neq\phi\}$.
Formally, $Syn^k(w) = \bigcup\limits_{i=1}^{k} Sy^{i}(w)$ for $i\in[1 \ldots k]$.

The example in Figure~\ref{fig:introsynEx} illustrates the computation.
The input Arabic word \RL{mA'}, denoted $w$, shares the gloss \cci{water} with the stem \RL{n.d.h}, denoted $s_1$, i.e. $s_1\in Sy^{1}(w)$.
%The input word \RL{mA'} shares the gloss \RL{n.d.h} through the gloss intersection \cci{water}. 
Next, the stem \RL{r^s^s}, denoted $s_2$, shares the gloss \cci{spray} with $s_1$, i.e. $s_2\in Sy^{1}(s1)\subset Sy^{2}(w)$.
Therefore, $Syn^2(w)$ relates the words \RL{mA'} and \RL{r^s^s}.

@@

Let ${\cal O} = \{ \mathit{isA}, \mathit{contains}\}$ be the set of atomic term 
predicates, and let ${\cal F} = \{ \pp, \ss, \xx, \PP, \GG, \AC\}$ be the 
set of morphological features where each entry is in turn a set of values for that feature.
Given a word $w$, a morphological feature $A\in{\cal F}$, 
a user defined constant feature value $CF\in A$, and an integer 
$k, 1\le k\le 7$, 
the following are morphology-based atomic terms (MAT), {\em terms} for short.
\begin{itemize}
  \item $a(w):= \exists m \in M(w). m=\langle p^k,s,x^l,P,G,C\rangle. r \circ CF$
where $\circ \in {\cal O}$, $r \in \{p^k,s,x^l,P,G,C\}$, and $r\in A$.
Informally, a solution vector of $w$ exists with
a feature containing or exactly matching $CF$.
The predicates $\mathit{isA}$ and $\mathit{contains}$ denote the exact match and contain criteria, respectively.
\item $a(w) := w \in Syn^k(CF), CF \in \ss$.
  Informally, this checks if $w$ is an extended synonym of a stem $CF$. 
\end{itemize}

@@

The match trees in Figure~\ref{fig:taskMRE} illustrate two matches of the expression computed by \framework.
The first match tree refers to the text 
\RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'wl}(Khalifa Tower next to the first intersection).
The second match tree refers to the text 
\RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building).
%The leaf nodes of the trees are entities and the edges and internal nodes are text, 
%morphology-based, 
%and word distance based relational entities. 
The leaf nodes of the match trees are matches to formulae and the internal nodes represent roots to sub-expression matches.
For instance, \RL{brj _hlyfT} in match 1 tree corresponds to the sub-expression $(P|N)+$.

>>>>

\subsection{The extended synonymy feature $Syn^k$}
\label{sec:synk}

\major{
Up to our knowledge, 
$Syn^k$ provides the first light Arabic wordnet based on the lexicon of Sarf.
}
The sets $E, A,$ and $L$ denote all English words, Arabic words, 
and Arabic lexicon words, respectively.
Recall that $\GG$ and $\ss$ denote the set of glosses and stems in the morphological analyzer, respectively.
We have $\GG \subset E$ and $\ss \subset L \subset A$. 
Function $\alpha: \ss \rightarrow 2^{\GG}$ maps Arabic stems to 
subsets of related English glosses, %; e.g. $g_{s} = \alpha(s)\subset 2^{\GG}$.
\major{
where $2^{\GG}$ denote the power set of $\GG$ which is the set of all subsets 
of $\GG$. 
}
Function $\gamma: L \rightarrow 2^{\ss}$ maps Arabic lexicon words to subsets 
of relevant Arabic stems. %; e.g. $s_{l} = \gamma(l)\subset 2^{\ss}$.

Given a word $w\in L$, 
$Sy(w)=\{u\mid u\in \ss \land\exists s\in \gamma(w)\land~\alpha(u)\cap\alpha(s)\neq\emptyset\}$
is the set of Arabic stems 
directly related to $w$ through the gloss map.
%
Let $Sy^{i}(w)$ denote stems related to $w$ using the gloss map of order $i$ recursively such that
$Sy^{1}(w) = Sy(w)$ and
$Sy^{i+1}(w)=\{u\mid u\in S\land\exists s\in Sy^{i}(w)\land~\alpha(u)\cap\alpha(s)\neq\phi\}$.
Formally, $Syn^k(w) = \bigcup\limits_{i=1}^{k} Sy^{i}(w)$ for $i\in[1 \ldots k]$.
%
\major{
The example in Figure~\ref{fig:introsynEx} illustrates the computation.
The input Arabic word \RL{mA'}, denoted $w$, shares the gloss \cci{water} with the stem \RL{n.d.h}, denoted $s_1$, i.e. $s_1\in Sy^{1}(w)$.
%The input word \RL{mA'} shares the gloss \RL{n.d.h} through the gloss intersection \cci{water}. 
Next, the stem \RL{r^s^s}, denoted $s_2$, shares the gloss \cci{spray} with $s_1$, i.e. $s_2\in Sy^{1}(s1)\subset Sy^{2}(w)$.
Therefore, $Syn^2(w)$ relates the words \RL{mA'} and \RL{r^s^s}.
}

@@

Let ${\cal O} = \{ \mathit{isA}, \mathit{contains}\}$ be the set of atomic term 
predicates, and let ${\cal F} = \{ \pp, \ss, \xx, \PP, \GG, \AC\}$ be the 
set of morphological features where each entry is in turn a set of values for that feature.
Given a word $w$, a morphological feature $A\in{\cal F}$, 
a user defined constant feature value $CF\in A$, and an integer 
$k, 1\le k\le 7$, 
the following are morphology-based atomic terms (MAT), {\em terms} for short.
\begin{itemize}
  \item $a(w):= \exists m \in M(w).~m=\langle p,s,x,P,G,C\rangle. r \circ CF$
where $\circ \in {\cal O}$, $r \in \{p,s,x,P,G,C\}$, and $r\in A$.
Informally, a solution vector of $w$ exists with
a feature containing or exactly matching $CF$.
The predicates $\mathit{isA}$ and $\mathit{contains}$ denote the exact match and contain criteria, respectively.
\item $a(w) := w \in Syn^k(CF), CF \in \ss$.
  Informally, this checks if $w$ is an extended synonym of a stem $CF$. 
\major{
  We limit $k$ to a maximum of $7$ since we practically noticed that 
  (1) values above $7$ introduce significant semantic noise and
  (2) the computation is expensive without a bound. 
}
\end{itemize}

@@

The match trees in Figure~\ref{fig:taskMRE} illustrate two matches of the expression computed by \framework.
The first match tree refers to the text 
\RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'wl}(Khalifa Tower next to the first intersection).
The second match tree refers to the text 
\RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building).
%The leaf nodes of the trees are entities and the edges and internal nodes are text, 
%morphology-based, 
%and word distance based relational entities. 
\major{
The leaf nodes of the match trees are matches to formulae and the internal nodes represent roots to sub-expression matches.
For instance, \RL{brj _hlyfT} in match 1 tree corresponds to the sub-expression $(P|N)+$.
}

### 2018-04-12 tex/sarf.tex
changes to answer, colors, and some reviewers addressed

\begin{table*}[bt]
\centering
\caption{Results of the hadith case study with Sarf.}
%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
\begin{tabular}{lp{.2cm}ccp{.2cm}ccp{.2cm}cc} %\cline{2-10}
 & & \multicolumn{2}{c}{Al Kafi} & & \multicolumn{2}{c}{Al Istibsar} & &\multicolumn{2}{c}{Ibn Hanbal} \\ \cline{3-10}
 & & Simple & Refined & & Simple & Refined & & Simple  & Refined \\\bottomrule  %\hline \hline
Word count & &\multicolumn{2}{c}{98,943} & & \multicolumn{2}{c}{103,835} & & \multicolumn{2}{c}{20,354} \\ 
 Names      & & 20,473 & 12,060  & &20,646 &  14,613& & 4,762 & 3,013\\
Names/Narrator& &2.48 & 1.97 & & 2.17 &  1.84& & 1.49 & 1.25 \\
Narrators & &3,080 & 2,623 & & 6,164 &  5,767& & 2,082 & 1,755 \\ 
Narrators/Chain & & 5.28 & 4.84 & & 4.98&  4.76 & & 4.8 &4.05 \\
Chains & & 583 & 542 &  & 1,238& 1,211& & 433 & 433 \\ 
Ignored names & & 11,801 & 6,400 &  & 5,845 & 3,348 & & 1,386 & 642 \\ \hline
Segmentation accuracy & & 85\% & 96\%& & 96\%& 96\%& & 92\%& 92\%\\ 
Chain accuracy & & 99\% & 99\%&  & 98\%& 99\%& & 99\% & 97\% \\ 
%Narrator accuracy & & & & & &  & & & \\ 
Narrator name accuracy & & 93\% & 91\%& &  92\%& 90\% & & 90\% & 90\% \\ \hline
Name false positives & & 28\% & 7\%&  & 9\%& 4\% & & 34\% & 4\% \\ \hline
Sarf running time (secs.)& & 2.97 & 1.32 & & 2.93 & 1.31 & & .25 & .096\\ \hline \hline
 & & Buckwalter& ElixirFM & & Buckwalter & ElixirFM & & Buckwalter & ElixirFM \\\bottomrule
Names      & & 18,354 & 19,201 & & 17,934 &  18,202 & & 5,021 & 5,117 \\
Names/Narrator& & 2.6 & 2.1 & & 2.47 &  2.07 & & 1.83 & 1.62\\
Narrators & & 3,210 & 2,981 & & 6,425 &  6,208 & & 2,311 & 1,965 \\ 
Narrators/Chain & & 5.57 & 5.32 & & 5.46 &  5.32 & & 5.12 & 5.01 \\
Chains & & 627 & 601 &  & 1,318 & 1,250 & & 487 & 452 \\ 
Ignored names & & 9,642 & 9,915 &  & 6,118 & 6,874 & & 1,812 & 1,904 \\ \hline
Segmentation accuracy & & 76\%& 84\%& & 85\%& 89 \%& & 84\%& 85\%\\ 
Chain accuracy & & 78\% & 87\%&  & 87\%& 90\%& & 85\% & 86\% \\ 
%Narrator accuracy & & & & & &  & & & \\ 
Narrator name accuracy & & 76\% & 86\%& & 84\%& 87\% & & 84\% & 84\% \\ \hline
Name false positives & & 29\% & 22\%&  & 14\%& 13\% & & 36\% & 32\% \\ \hline
Running time (secs.)& & 6.65 & 12$\times$60& &4.55 & 14$\times$60 & & .66 & 6$\times$60 
\end{tabular}
\normalsize
\label{t:hadithresallresults}
\end{table*}

@@

In this paper we o

>>>>

\begin{table*}[bt]
\centering
\caption{Results of the hadith case study with Sarf.}
%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
\begin{tabular}{lp{.2cm}ccp{.2cm}ccp{.2cm}cc} %\cline{2-10}
 & & \multicolumn{2}{c}{Al Kafi} & & \multicolumn{2}{c}{Al Istibsar} & &\multicolumn{2}{c}{Ibn Hanbal} \\ \cline{3-10}
 & & Simple & Refined & & Simple & Refined & & Simple  & Refined \\\bottomrule  %\hline \hline
Word count & &\multicolumn{2}{c}{98,943} & & \multicolumn{2}{c}{103,835} & & \multicolumn{2}{c}{20,354} \\ 
 Names      & & 20,473 & 12,060  & &20,646 &  14,613& & 4,762 & 3,013\\
Names/Narrator& &2.48 & 1.97 & & 2.17 &  1.84& & 1.49 & 1.25 \\
Narrators & &3,080 & 2,623 & & 6,164 &  5,767& & 2,082 & 1,755 \\ 
Narrators/Chain & & 5.28 & 4.84 & & 4.98&  4.76 & & 4.8 &4.05 \\
Chains & & 583 & 542 &  & 1,238& 1,211& & 433 & 433 \\ 
Ignored names & & 11,801 & 6,400 &  & 5,845 & 3,348 & & 1,386 & 642 \\ \hline
Segmentation accuracy & & 85\% & 96\%& & 96\%& 96\%& & 92\%& 92\%\\ 
Chain accuracy & & 99\% & 99\%&  & 98\%& 99\%& & 99\% & 97\% \\ 
%Narrator accuracy & & & & & &  & & & \\ 
Narrator name accuracy & & 93\% & 91\%& &  92\%& 90\% & & 90\% & 90\% \\ \hline
Name false positives & & 28\% & 7\%&  & 9\%& 4\% & & 34\% & 4\% \\ \hline
Sarf running time (secs.)& & 2.97 & 1.32 & & 2.93 & 1.31 & & .25 & .096\\ \hline \hline
 & & Buckwalter& ElixirFM & & Buckwalter & ElixirFM & & Buckwalter & ElixirFM \\\bottomrule
Names      & & 18,354 & 19,201 & & 17,934 &  18,202 & & 5,021 & 5,117 \\
Names/Narrator& & 2.6 & 2.1 & & 2.47 &  2.07 & & 1.83 & 1.62\\
Narrators & & 3,210 & 2,981 & & 6,425 &  6,208 & & 2,311 & 1,965 \\ 
Narrators/Chain & & 5.57 & 5.32 & & 5.46 &  5.32 & & 5.12 & 5.01 \\
Chains & & 627 & 601 &  & 1,318 & 1,250 & & 487 & 452 \\ 
Ignored names & & 9,642 & 9,915 &  & 6,118 & 6,874 & & 1,812 & 1,904 \\ \hline
Segmentation accuracy & & 76\%& 84\%& & 85\%& 89 \%& & 84\%& 85\%\\ 
Chain accuracy & & 78\% & 87\%&  & 87\%& 90\%& & 85\% & 86\% \\ 
%Narrator accuracy & & & & & &  & & & \\ 
Narrator name accuracy & & 76\% & 86\%& & 84\%& 87\% & & 84\% & 84\% \\ \hline
Name false positives & & 29\% & 22\%&  & 14\%& 13\% & & 36\% & 32\% \\ \hline
Running time (secs.)& & 6.65 & 12$\times$60& &4.55 & 2.3$\times$3600 & & .66 & 29.2$\times$60 
\end{tabular}
\normalsize
\label{t:hadithresallresults}
\end{table*}

@@
